---
title: Visualize and analyse IMDB ratings with R
author: Martijn van Vreeden
date: '2017-09-28'
slug: visualize-and-analyse-imdb-ratings-with-r
categories:
  - digital me
  - imdb
tags:
  - imdb
  - me
  - r
  - ggplot2
description: 'I will explore the data to learn about my movie preferences. I will be using some R packages to structure the data properly and other packages to build plots' 
content_summary: "In this post I will explore the data to learn about my movie preferences. I will be using some R packages to structure the data properly and other packages to build plots."
---

_This post is part of a series of posts to [analyze the digital me](/meet-the-digital-me/)._

# Visualize and analyse IMDB ratings with R
In [my previous post](/getting-my-imdb-ratings-with-r-and-rvest/) in this series I shared how I collect and store my personal ratings data from IMDB and enrich it with a few important variables about the movie.

In this post I will explore the data to learn about my movie preferences. I will be using some R packages to structure the data properly and other packages to build plots:

* [plyr](https://cran.r-project.org/web/packages/plyr/index.html)
* [dplyr](https://cran.r-project.org/web/packages/dplyr/index.html)
* [tidyr](https://cran.r-project.org/web/packages/tidyr/index.html)
* [ggrepl](https://cran.r-project.org/web/packages/ggrepel/index.html)
* [lubridate](https://cran.r-project.org/web/packages/lubridate/index.html)
* [ggplot2](https://ggplot2.tidyverse.org/)

I'm sure my code can be optimized to use less packages or run faster in general. Feel free to reach out to me if you have suggestions on how to do this. I'm learning as I'm going.

```{r start, echo=FALSE, message=FALSE, warning=FALSE}
library('ggplot2')
library('plyr')
library('dplyr')
library('plotly')
library('RMySQL')
library('tidyr')
library('ggrepel')
library('lubridate')
library('ggthemes')
library('scales')
read.csv("dfs/imdb.csv") -> data_imdb
```

## Load data.frame with IMDB rating data

I have loaded my data.frame with my ratings, IMDB ratings and general movie info to analyse. You can see how I collected this data in [my previous post](/getting-my-imdb-ratings-with-r-and-rvest/).

```{r table}
head(data_imdb)

```

Let's first get some summary data to see what is in my data.frame.

```{r summary, warning=FALSE}
#Number of movies in dataframe
nrow(data_imdb)

# Average IMDB from all users on movies
round(mean(data_imdb$imdb_rating),2)

#Average rating from me on the same movies
round(mean(data_imdb$myRating),2)
```

What we can immediately see is that I'm not easy to please. My average rating is a lot lower than the one from other IMDB raters.

```{r summary2, warning=FALSE}
#My personal top 10 of movies I've rated
head(arrange(data_imdb[,c("title", "myRating")],desc(myRating)),10)

#IMDB users top 10 of movies I've rated
head(arrange(data_imdb[,c("title", "imdb_rating")],desc(imdb_rating)),10)
```

From the looks of the top10's my taste isn't totally aligned with the IMDB crowd either. A lot of my favorites are not in the IMDB list.

## Comparing my ratings to those of all IMDB users

Let's dive into those differences a bit more. 
In order to do this I will create an additional column in my dataframe to show the differences in ratings per movie. I will use the data to show a top10 of biggest positive difference in ratings (I like them a lot more than IMDB users) and the other way around (movies I don't really like that IMDB users do).

```{r differences, warning=FALSE}
data_imdb$dif_rating <- data_imdb$myRating - data_imdb$imdb_rating

#Biggest differences I like a lot more than IMDB users
head(arrange(data_imdb[,c("title", "dif_rating")],desc(dif_rating)),10)

#Biggest differences IMDB users like a lot more than I do
tail(arrange(data_imdb[,c("title", "dif_rating")],desc(dif_rating)),10)
```

Clearly I don't get the appeal of the [Thor movies](https://www.imdb.com/list/ls031343822/), [The Expandables 2](https://www.imdb.com/title/tt1764651/) and [13](https://www.imdb.com/title/tt0798817/), but I somehow thought [Hobo with a shotgun](https://www.imdb.com/title/tt1640459/) was a masterpiece when nobody else seemed to get that movie.

## Ratings per decade of movies

I want to learn more about my general preferences, not on an individual movie level, but grouped into segments. Am I a big fan of movies from a particular decade?

### Number of movies I watched from each decade

I'm visualising this data with the [ggplot2 package](https://ggplot2.tidyverse.org/). A simple way to make graphs and plots based on dataframes. There a lots of resources about this package to learn how it works.

```{r decade, warning=FALSE}
theme_data <- theme_light(base_size = 13) + theme(plot.margin=unit(c(0.5,1,1.5,1.2),"cm")) 
color_bar <- "#999999"
color_fill <- "#5c85d6"
#spread of movie release decades
(data_imdb$year %/% 10) * 10 -> data_imdb$decade
ggplot(data = data_imdb, aes(x = decade))+
  geom_bar(stat = 'count', fill=color_fill, color = color_bar) +
  scale_x_continuous(breaks = c(1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
  scale_y_continuous(expand = c(0, 0),breaks = seq(0, 325, 25), limits = c(0,325)) +
  geom_text(stat='count',aes(label=..count..),vjust=-0.5) +
  ylab("Number of movies watched") +
  theme_data
```
Most of the movies I have reviewed are from the past 20 years. I still have to rate my first forties of fifties movie on IMDB...

### My ratings of movies per decade
```{r my_ratings_decade, warning=FALSE}
ggplot(data = data_imdb, aes(x = decade, y = myRating))+
  geom_bar(stat = "summary", fun = "mean", fill = color_fill, color = color_bar) + 
  ylab("Average rating") +
  theme_data +
  scale_x_continuous(breaks = c(1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010)) +
  scale_y_continuous(limits = c(0,10), breaks = c(0:10)) +
  stat_summary(aes(label=round(..y..,2)), fun.y=mean, geom="text", vjust = -0.75)
```

### Difference in ratings between me and all IMDB users

```{r rating_comparison_decade, warning=FALSE}
ggplot(data = data_imdb, aes(x = decade, y = dif_rating))+
  geom_bar(stat = "summary", fun = "mean", fill=color_fill, color = color_bar)+
  ylab("Difference in average rating") +
  theme_data +
  stat_summary(aes(label=round(..y..,2)), fun=mean, geom="text", vjust = -0.75) +
  scale_y_continuous(limits = c(-2,2), breaks = c(-2,-1,0,1,2))
```

It turns out I'm not giving very high ratings in comparison to all IMDB users, regardless of the decade the movie is from. Based on my ratings per decade and the difference between me and IMDB users, it would make sense to enjoy some more movies from the nineties and older than the nineties. I'm really tough on movies from the past 20 years in comparison to other IMDB raters (-0.37 and -0.43). 

On the other hand, I'm watching a lot of movies that were released in the past 20 years. Maybe I should just pick them more carefully. Let's find out if there is a bigger range of ratings I give to movies from the past 20 years than other years.

### Distribution of my ratings per decade

I will be using a boxplot to visualise the distribution of my ratings per decade. If you are not sure how to read a boxplot, this is [a great explanation about reading boxplots](https://flowingdata.com/2008/02/15/how-to-read-and-use-a-box-and-whisker-plot/) for you to check out.

```{r rating_distribution_decade, warning=FALSE}
ggplot(data_imdb, aes(x = decade, y = myRating, group = decade)) +
  geom_boxplot() +
  ylab("My average rating") +
  scale_x_discrete(limit = c(1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010)) + 
  theme_data
```
The boxplots can be a little oddly shaped, because I can only rate with whole values (only 7 or 8, not 7.5). A full quartile can consist of one value (ie 6 or 7). The median can also be the same value as the upper quartile or lower quartile. 

In this boxplot there are not a lot of outliers for the past 20 years. This means I'm consistently watching movies that I rate 6 or 7 stars. It is a lot more positive for the eighties and nineties. I'm also rating movies from those times with lower stars, but that isn't as common. I guess I should be watching more movies older than 20 years.

In my next post I will analyse this dataframe further. I still have to learn more about my favorite actors, directors and genres. I hope to get some good recommendations out of this for movie night.
