---
title: Visualize and analyse IMDB ratings with R (part 2)
author: Martijn van Vreeden
date: '2017-10-07'
slug: visualize-and-analyse-imdb-ratings-with-r-part-2
categories:
  - digital me
  - imdb
tags:
  - ggplot2
  - imdb
  - me
  - r
  - webscraping
  - rvest
description: 'I will explore the data to learn about my movie preferences. I will be using some R packages to structure the data properly and other packages to build plots'
content_summary: "In my previous post about analysing IMDB ratings in this series I explored some of the data I collected about my movie preferences, also compared to the ratings other IMDB raters. In this post I will dig a little deeper to learn more about my own personal movie preferences."
---

# Visualize and analyse IMDB ratings with R (part 2)

_This post is part of a series of posts to [analyse the digital me](/meet-the-digital-me/)._

In [my previous post about analysing IMDB ratings](/visualize-and-analyse-imdb-ratings-with-r/) in this series I explored some of the data I collected about my movie preferences, also compared to the ratings other IMDB raters.

In this post I will dig a little deeper to learn more about my own personal movie preferences. 

```{r start, echo=FALSE, message=FALSE, warning=FALSE}
library('ggplot2')
library('plyr')
library('dplyr')
library('plotly')
library('RMySQL')
library('tidyr')
library('ggrepel')
library('lubridate')
read.csv("dfs/imdb.csv") -> data_imdb
```

## Does popularity of a movie impact my rating?

Some movies are very mainstream, other are more niche or an "acquired" taste. The popularity of a movie can be determined in IMDB by the number of reviews a movie has received. Let's first see how the number of IMDB reviews relates to the average IMDB rating. We can do that by building a sctaterplot.

```{r IMDB_ratings, warning=FALSE,message = FALSE}
theme_data <- theme_light(base_size = 13) + theme(plot.margin=unit(c(0.5,1,1.5,1.2),"cm")) 
color_bar <- "#999999"
color_fill <- "#5c85d6"
#spread of movie release decades
ggplot(data_imdb, aes(x = imdb_ratings, y = imdb_rating)) +
  geom_point() +
  #scale_x_discrete(limit = c(1930, 1940, 1950, 1960, 1970, 1980, 1990, 2000, 2010)) + 
  scale_y_discrete(limit = c(3:10)) + 
  ylab("IMDB average rating") +
  xlab("Number of IMDB ratings") +   
  theme_data
```

In this scatterplot you can see that most of the movies are clustered around a low number of reviews and are receiving average ratings between 6 and 7. It's interesting to see that the movies with most IMDB ratings are all highly rated. Let's check out which movies these are.

```{r IMDB_top10_ratings, warning=FALSE,message = FALSE}
head(arrange(data_imdb[,c("title", "imdb_ratings", "imdb_rating")],desc(imdb_ratings)),10)
```

So Shawshank Redemption and Dark Knight are the most reviewed movies I have rated myself. And they are both highly rated.

### Understanding the probability of number of ratings to impact my rating

The probability can be calculated with a p-value.

```{r correlation-matrix, warning=FALSE,message = FALSE}
cor(as.numeric(data_imdb$imdb_ratings), as.numeric(data_imdb$myRating))
```

As you can see, the p-value is `r cor(as.numeric(data_imdb$imdb_ratings), as.numeric(data_imdb$myRating))`. This means that the number of ratings on IMDB will have a 45% chance of having no effect on my own rating. So there is a pretty high probability that the number of ratings will have an effect on my rating (52%). 

## My favorite actors

I always think I know which actors I like and put their movies on the top of my watchlist. Let's check out if I actually rate the movies my favorite actors with the highest scores too.

For each movies I rate, there are a maximum of 15 actors added to the dataset. I have added them all to one column, pipe separated. I will use the [data.table](https://cran.r-project.org/web/packages/data.table/vignettes/datatable-intro.html) and [stringr](https://cran.r-project.org/web/packages/stringr/vignettes/stringr.html) package to split the data in the cast column and learn more about the actors in the movies I reviewed. I will create a new dataframe with a count of the the actors (unique_actor_count) and find out how many actors are in this dataframe (unique_actors). After that, I will create a new dataframe with my top 10 most watched actors and visualise it in a barchart for easy interpretation.

```{r IMDB_actor_count, warning=FALSE,message = FALSE}
library('data.table')
library('stringr')
as.character(data_imdb$imdb_cast) -> data_imdb$imdb_cast
data.frame(table(unlist(strsplit(data_imdb$imdb_cast, "[|]")))) -> unique_actor_count
```

```{r IMDB_actor_count_hidden, echo=FALSE, message=FALSE, warning=FALSE}
 unique_actor_count <-  unique_actor_count[!unique_actor_count$Var1 =="Rest of cast listed alphabetically:",]
 unique_actor_count <-  unique_actor_count[!unique_actor_count$Var1 =="",]
 unique_actor_count = unique_actor_count[-1,]
```

```{r IMDB_actor_count1, warning=FALSE,message = FALSE}
nrow(unique_actor_count) -> unique_actors
head(arrange(unique_actor_count,desc(Freq)),10) -> unique_actor_count_top10

ggplot(data = unique_actor_count_top10, aes(x = Var1, y = Freq)) +
  geom_bar(stat = "identity", fill=color_fill, color = color_bar)+
  coord_flip() +
  geom_text(aes(label = Freq), position=position_dodge(width=1), hjust = -0.5) + 
  theme(axis.text.x = element_text(angle = 90, hjust = 1)) +
  ylab("Number of movies watched featuring actor") +
  xlab("") +
  theme_data

```

It turns out more than 8000 actors have made it to my dataframe and the one I have viewed more than others was Samuel L. Jackson. I'm surprised to see only male actors made it to my top 10. Not sure how that happened...

### Average rating of my most popular actors

Next, I want to know if my most popular actors are in highly rated movies. I'll stick to my top 100 most popular actors and visualise these in a scatterplot. This means I will select the top100 actors, merge them with the dataframe that has all the movies rating data in there and create a scatterplot where I label the points with the actor names using the [ggrepel package](https://cran.r-project.org/web/packages/ggrepel/index.html).

```{r IMDB_actor_ratings, warning=FALSE, message = FALSE}
library('ggrepel')
head(arrange(unique_actor_count,desc(Freq)),100) -> unique_actor_count_top100
unique_actor_count_top100$Var1 -> actor_group
df_mean_actor <- data.frame()
for(v in actor_group){
  data_imdb %>%
    filter(str_detect(imdb_cast, v)) %>%
    select(myRating) -> df9
  round(mean(df9$myRating),2) ->mean_actor
  df10 <- data.table("actor" = v, "mean" = mean_actor)
  df_mean_actor<-rbind(df_mean_actor,df10)
}

actor_overview = merge(unique_actor_count_top100, df_mean_actor, by.x=c("Var1"), by.y=c("actor"))

ggplot(data = actor_overview, aes(x = mean, y = Freq)) +
  geom_point() +
  geom_text_repel(aes(label=ifelse(mean>7.5,as.character(Var1),'')), color = "forestgreen", box.padding = unit(0.8, "lines")) +
  geom_text_repel(aes(label=ifelse(mean<5.8,as.character(Var1),'')), color = "red", box.padding = unit(2.2, "lines")) +
  ylab("Number of movies watched featuring actor") +
  xlab("Average rating") +
  theme_data
```
This visualisation gives me some clear insights:

* [Nicolas Cage](https://www.imdb.com/name/nm0000115/) and [Sylvester Stallone](https://www.imdb.com/name/nm0000230/) movies should be blacklisted for now. I watched several of them, but I rate them poorly
* I give [Al Pacino](https://www.imdb.com/name/nm0000199/) movies high ratings and I have seen a lot of his movies. Safe to watch his other movies too.
* [Michael Caine](https://www.imdb.com/name/nm0000323/) is the hidden gem. I have hardly watched his movies, but I loved them. Time to look for some more of his movies!
* I haven't watched many movies with [Jason Biggs](https://www.imdb.com/name/nm0004755/) or [Ray Stevenson](https://www.imdb.com/name/nm0829032/) and I shouldn't start any time soon :)

## My favorite movie genres

Just like my favorite actors, I can also learn more about favorite genres. The data is setup in the same way, so for now I will go straight to the visualisations.

```{r unique_genre_count, warning=FALSE}
as.character(data_imdb$imdb_genres) -> data_imdb$imdb_genres
data.frame(table(unlist(strsplit(data_imdb$imdb_genres, "[|]")))) -> unique_genres_count
c("Addcontentadvisoryforparents", "MPAA", "Seeallcertifications", "Viewcontentadvisory") -> genre_dirt
filter(unique_genres_count, !grepl(paste(genre_dirt, collapse="|"), Var1)) -> unique_genres_count
nrow(unique_genres_count)
```
Just like actors, it is possible for a movie to have multiple genres. I have made them unique and can now see that I only have `r nrow(unique_genres_count)` genres in the dataset. 

Let's see what the average rating is per genre and the number of movies I have watched by genre.

```{r genre_rating, warning=FALSE}
head(arrange(unique_genres_count,desc(Freq)),10) -> unique_genres_count_top10

unique_genres_count$Var1 -> genre_group

df_mean_genre <- data.frame()
for(v in genre_group){
  data_imdb %>%
    filter(str_detect(imdb_genres, v)) %>%
    select(myRating) -> df7
    round(mean(df7$myRating),2) ->mean_genre
    df8 <- data.table("genre" = v, "mean" = mean_genre)
    df_mean_genre<-rbind(df_mean_genre,df8)
}

genre_overview = merge(unique_genres_count, df_mean_genre, by.x=c("Var1"), by.y=c("genre"))

ggplot(data = genre_overview, aes(x = Var1, y = mean, fill = Freq)) +
  geom_bar(stat = "identity")+
  geom_text(aes(label = mean), size = 3, position=position_dodge(width=1), vjust = -1) + 
  scale_y_continuous(expand = c(0, 0),breaks = seq(0, 10, 1), limits = c(0,8)) +  
  ylab("Average rating") +
  xlab("Genre") +
  theme_data +  
  theme(axis.text.x = element_text(angle = 90, hjust = 1))
```

I really enjoy a great animation movie (brings out the child in me), but I hardly watch them. Horror is not my preferred genre and that shows in the number of horror movies I have seen and how I rate them.
